---
title: "Graphs with ggplot"
author: "Ken Mwai"
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "sidebartab"
date: "Friday, January 16, 2015"
framework   : io2012
fontsize: 11pt 
---

## What is ggplot2?

- An implementation of the __Grammar of Graphics__ by Leland Wilkinson
- Grammar of graphics represents and abstraction of graphics ideas/objects
- Think “verb”, “noun”, “adjective” for graphics
- Allows for a “theory” of graphics on which to build new graphics and graphics objects


## Basic Components of a ggplot2 Plot
- A _data frame_
- _aesthetic mappings_: how data are mapped to color, size 
- _geoms_: geometric objects like points, lines, shapes. 
- _facets_: for conditional plots. 
- _stats_: statistical transformations like binning, quantiles, smoothing. 
- _scales_: what scale an aesthetic map uses (example: male = red, female = blue). 
- _coordinate system_ 

## Basic Quick Plot

- Load the maltreat dataset 

```{r , echo=FALSE , results='hide'}
library(knitr)
opts_chunk$set(out.width='200px', dpi=200 , fig.align ='center', warning=FALSE , message=F)
maltreat<-read.csv("maltreat.csv")
maltreat<-subset(maltreat, sex!="")
```

```{r , warning=FALSE ,echo=T}
library(ggplot2)
qplot(x = weight , data = maltreat )
```


## Basic Quick scatter plot

```{r , warning=FALSE}
library(ggplot2)
qplot(x = weight  , y=heartrate, data = maltreat )
```


## Build your graph in layers
  Basic Components of a ggplot2 Plot
  ---------------------------------
      - A _data frame_
      - _aesthetic mappings_: how data are mapped to color, size 
      - _geoms_: geometric objects like points, lines, shapes. 
      - _facets_: for conditional plots. 
      - _stats_: statistical transformations like binning, quantiles, smoothing. 
      - _scales_: what scale an aesthetic map uses (example: m = red, f = blue). 
      - _coordinate system_ 
      
* * *      

 - ggplot(df -- **Data**, aes(x, y, <other aesthetics>) -- **what to map**) \newline
 + geoms_ **what type of graph**
 
## Types of Geoms
   - geom_bar() for **bar graphs**
   - geom_histogram() for **histograms** -- diff??
   - geom_line() for **line graphs** 
   - geom_boxplot()  **Box and whiskers plot**
   - geom_errorbar() **Error bars.**
   
   
## Example - Box Plot
```{r}
boxplot(maltreat$weight)
```

* * *
  - call library(ggplot2)
  - ggplot(data = maltreat)
  - ggplot(data = maltreat , aes(weight))
  
```{r}
library(ggplot2)
g <- ggplot(data = maltreat , aes(weight))
```

* * *
 - ggplot(data = maltreat , aes(weight)) + geom_boxplot()
 
```{r}
library(ggplot2)
ggplot(data = maltreat , aes(x=weight , y=weight)) +
  geom_boxplot()
```


## Boxplot of weight and sex 
```{r}
boxplot(maltreat$weight~maltreat$sex,col="grey")
```

 - Prepare in ggplot2

* * *
- This is it


})
```{r}
library(ggplot2)
ggplot(data = maltreat , aes(x=sex , y=weight)) + 
  geom_boxplot(fill = "red")
```

* * *
 
```{r}
library(ggplot2)
ggplot(data = maltreat , aes(x=sex ,
                    y=weight ,fill=paying)) + 
  geom_boxplot()
```

## Histogram of weight 
```{r}
hist(maltreat$weight, col="grey")
```

 - Prepare in ggplot2
 
 
 * * *
- This is the Histogram
 
```{r}
library(ggplot2)
ggplot(data = maltreat , aes(weight)) + geom_histogram()
```


## Scatter of weight  vs heartrate
```{r}
plot(maltreat$weight,maltreat$heartrate, pch=19,col="blue")
```

 - Prepare in ggplot2
 
 
 * * *
- This is the scatter
 
```{r}
library(ggplot2)
ggplot(data = maltreat , aes(x=weight , y=heartrate)) + 
  geom_point()
```


## Bar plot of sex
```{r}
barplot(table(maltreat$sex))
```

 - Prepare in ggplot2
 
 
 * * *
- This is the barplot \\o/
 
```{r}
library(ggplot2)
ggplot(data = maltreat , aes(sex)) + geom_bar(fill='grey') +
  theme_bw() + ggtitle("Barplot of gender")
```


## A detailed boxplot

- draw a boxplot of resprate by gender 
- filled by paying
- has a titlle
- has an xlabel
- has a y label
- saved in png

* * *
Like This 

```{r, warning=FALSE,echo=FALSE}
library(ggplot2)
maltreat$paying<- factor(maltreat$paying,
levels = c("","Yes"),
labels = c("No", "Yes")) 
ggplot(data = maltreat , aes(x=sex , y=resprate ,fill=paying)) +
  geom_boxplot(position=position_dodge(.8)) + 
  ggtitle("A box blot of Resprate by gender") +
  xlab("Gender") + ylab("Resprate") +
  theme_bw() 
```

* * *
```{r, warning=FALSE,echo=T , size="tiny"}
library(ggplot2)
maltreat$paying<- factor(maltreat$paying,
levels = c("","Yes"),
labels = c("No", "Yes")) 
ggplot(data = maltreat , aes(x=sex , y=resprate ,fill=paying)) +
  geom_boxplot(position=position_dodge(.8)) + 
  ggtitle("A box blot of Resprate by gender") +
  xlab("Gender") + ylab("Resprate") +
  theme_bw() 
```
