---
title: "Survival Analysis using R"
author: "Phillip Ayieko"
header-includes:
   - \usepackage{multicol}
   - \usepackage{graphicx}
output:
  beamer_presentation:
    theme: "CambridgeUS"
    colortheme: "sidebartab"
    keep_tex: true
framework   : io2012
---

## Survival Topics
- Survival Analysis concepts
- Common survival analysis techniques
- Presenting results
- Both theory and practice
    - Definitions
    - censoring
    - Kaplan-Meier
    - Log-rank
    - Cox proportional hazards
    - R  statistical software (open source: http://cran.r-project.org/) 

## Introduction to survival analysis
- Survival Analysis is referred to statistical methods for analyzing survival/ time to event data
- Events could be
    - Recurrence of disease, death etc
- Examples
    - Time between surgery and relapse
- Data is usually censored
    - Example-study ends before all the patients dies
    
## Objective of survival analysis

- Estimate probability that an individual surpasses some time-to-event for a group of individuals.
- Compare time-to-event between two or more groups.
- Assess the relationship of covariates to time-to-event.

## Types of survival data

- When the time of event is known, then we have complete information e.g. rat number 4 died on 3rd day. This is called uncensored data
- When the occurrences of the event are not known then we have incomplete information e.g. those rats who escaped. This is called censored data
- Follow-up time, calculated from
    - exit time when the outcome is experienced, or last-seen time
    - entry time when individual enters study
    
## Two reasons for censoring   
- During investigation we have incomplete information due to LOSS TO FOLLOW UP
- When the investigation has come to an end and the experimental units are still alive. This is censoring due to WITHDRAWAL ALIVE
    
## Forms/ Types of censoring    
- Right censoring
- Left censoring
- Type I censoring
- Type II censoring
- Random censoring
- Interval censoring
- Informative and non-informative censoring
    - Most common is right censored
    
## Censoring
- How could we account for censoring?
    - Ignore it and say event occurred at time of censoring
        - Incorrect because this is almost certainly not true
    - Remove patient from analysis
        - Potential bias and loss of power
    - Survival analysis
- The objective is to estimate the survival distribution of patients in the presence of censoring

## Functions describing survival times

- T is a random variable representing survival time - Can be described by a probability distribution

    - 1.Cumulative distribution function
    - 2.Survival function
    - 3.Probability density function
    - 4.Hazard function
    - 5.Cumulative hazard function
    
## Functions of T
- Let **T** denote the survival time
- $t \equiv$ a specific point in time.

1. **Survival function**
  - $s(t)=p(surving\_longer_than\_time\_t)$

- Survival function: probability survival time is greater than a specific value $S(t)=P(T>t)$

- $S=\frac{number\_of_patients\_surviving\_longer\_than\_t}{total\_number\_of\_patients\_in\_the\_study}$

## Mathematical Definitions
 - Distribution function
$$
 F(t)=Pr(T\leq t)
$$
 
 - Density Function
$$
f(t)=\frac{\partial{F(t)}}{\partial{t}}
$$ 
 
 - Survival Function
$$
S(t)=Pr(T>t)=1-F(t)
$$ 
 

## Mathematical Definitions
- Hazard function:
$$
h(t)=\lim_{\Delta t \to 0} \frac{Pr(t\leq T \leq t + \Delta t|T\geq t)}{\Delta t} = \frac{f(t)}{S(t)}
$$
- Cumulative hazard

$$
H(t)=\int_{0}^{t} h(u) du
$$

## Statistical Inference

- Has two components
    - Estimation of survival function
    - Testing of hypothesis of survival functions
- There are two methods of approaching them
    - Parametric approach
    - Non parametric approach
- A combination of the 2 approaches gives the regression (i.e. estimation and testing of the hypothesis)

##  KAPLAN-MEIER ESTIMATOR

## Kaplan-Meier Estimate of S(t)

- Kaplan-Meier estimator of survival is a nonparametric method of inference concerning the survivor function $S = Pr(T > t)$
- Rank the survival times as $t_1 \leq t_2 \leq ... \leq t_n$
- Number of individuals at risk before $t_i \equiv n_i$
- Number of individuals with failure time $t_i \equiv d_i$
- Estimated hazard function at $t_i$: $h_i=\frac{d_i}{n_i}$

- Formula
$$
S(t) = \prod_{t_i \leq t} \frac{n_i - d_i}{n_i}
$$


## Kaplan-Meier Estimate of S(t)
$$
var[s(t)]\approx [s(t)]^2\sum \frac{d_j}{n_j(n_j-d_j)}
$$

- And the 95 % CI is given by

$$
s(t)\pm 1.96 \sqrt{var(s(t))}
$$

## Kaplan-Meier Curve

```{r, echo=FALSE ,message=FALSE, warning=FALSE}
 library(survival)
 hmohiv<-read.table("hmohiv.csv", sep=",", header = TRUE) 

 attach(hmohiv)
 mini<-hmohiv[ID<=5,]
 attach(mini)
mini.surv <- survfit(Surv(time, censor)~ 1, conf.type="none")
plot(mini.surv, xlab="Time", ylab="Survival Probability")
```


## Functions of interest in R
- Survival object: **Surv**
- Kaplan-Meier estimates: **survfit**
- The log-rank test: **survdiff**
- The Cox proportional hazards model: **coxph**
- The Accelerated failure time model: **survreg**
- Relevant R packages: **survival, survcomp, HMISC, Design, MASS**

##Example
- Event = time to relapse
- Data:
    - 10, 20+, 35, 40+, 50+, 55, 70+, 71+, 80, 90+
- Before complex functions may be performed, the data has to be put into the proper format: a **survival object**
- In R we use **Surv(time, status)**
- *time* is a vector of event time, and *status* is a vector of indicator (denoting if the event was observed or censored

##R codes for Kaplan-Meier

```{r , echo=TRUE , warning=FALSE ,message=FALSE}
library(survival)

time <- c(10,20,35,40,50,55,70,71,80,90)
status <- c(1,0,1,0,0,1,0,0,1,0)
data <- data.frame(time,status)

survobj <- with(data,Surv(time,status))

fit <- survfit(survobj~1, data=data)
fit
summary(fit)
attributes(fit)

```

***

```{r, echo=TRUE}
plot(fit, xlab="time to relapse (months)", ylab="Survival Function", yscale=100, main="Survival Distribution")
```

## Aalen-Nelson Estimate of S(t)
- Uses the cumulative hazard function
$$
H(t) = - \log s(t) = \sum_{t_j\leq t}\frac{d_j}{n_j}
$$

$$
s(t)=e^{-\int_{0}^{t} h(u)d(u)} =e^{-H(t)}
$$

- In R, to estimate S(t), using output from *survfit()*

```{r}
#survobj <- Surv(time, status)
#fit1 <- summary(survfit(survobj~1))
# S.hat <- fit1$surv
# H.hat <- −log(S.hat)
#a<- survfit(coxph(Surv(time,status)~1), type="aalen")
#summary(a)
#basehaz(coxph(Surv(time,status)~1,data=data))
```

## LOG-RANK TEST

## LOG-RANK TEST
- Use the Log-Rank Test to compare the survival functions of two samples/ treatment groups.

  - $H_0$: The two survival functions are the equivalent
  - $H_a$: The two survival functions are different

- Goal: Test whether two groups differ in population survival functions.

## Log-Rank Test
 - Notation:
   - $t_{i} \equiv$ Time of the ith failure time (across groups)
   - $d_{1i}  \equiv$  number of failures for trt 1 at time $t_{(i)}$
   - $d_{2i}  \equiv$  Number of failures for trt 2 at time $t_{(i)}$
   - $n_{1i} \equiv$  Number at risk prior for trt 1 prior to time $t_{(i)}$
   - $n_{2i}$  Number at risk prior for trt 2 prior to time $t_{(i)}$
  
$d_i = d_{1i}+ d_{2i}$  and    $n_i=n_{1i}+n_{2i}$
$$
v_{li} =\frac{ n_{1i}n_{2i}d_i(n_i-d_i) }{n_{i}^{2} (n_i-1)}
$$ 

$$
e_{1i} = \frac {n_{1i} d_{1i}}{n_i}
$$

$$
O_1-E_1 =\sum_i (d_{1i}-e_{1i})
$$ 

$$
V_1=\sum_i v_{li}
$$

## R codes
```{r , eval=FALSE}
survdiff(formula, rho=0)
```
- *formula* is a survival object against a categorical covariate variable.
- *rho* is a scalar parameter that controls the type of test.
- With default *rho=0*, this is the log-rank or Mantel-Haenszel test.
- With *rho=1*, it is equivalent to the Peto and Peto modi???cation of the Gehan-Wilcoxon test.
```{r , eval=FALSE}
  >survdiff(survobj~status, rho=0)
```


##  Cox Proportional Hazard Model
        
        
## Cox Proportional Hazard Model
- CPHM is a technique for investigating the relationship between survival time and independent variables 
- The most popular model for survival analysis:
- Goal: Compare two or more groups (treatments), adjusting for other risk factors on survival times (like Multiple regression)
- p explanatory variables (including dummy variables)

## Cox regression for time to event

- Useful when incidence rates vary over time

- hazards in both exposed and unexposed are allowed to vary but their ratio is assumed to be constant - this is the proportional hazards assumption. Thus the Cox model are also known as the proportional hazards model

- Cox model is fitted by creating risk sets of individuals still at risk at the time of each failure

##Cox Proportional Hazard Model
 - The semi-parametric model given by
$$
h_i(t)=h_0(t)\exp(\beta_1 x_{i1} + \beta_2 x_{i2} + ... + \beta_k x_{ik})
$$

- $h_0(t)$ is the baseline hazard function. This is the function when all the covariates equal to zero
- The hazard ratio is independent of time t. 
$$
HR_{i,j}=\frac{h_i(t)}{h_i(t)} +\frac{\lambda_0(t)e^{\beta_1 x_{i1} + ... + \beta_k x_{ik}}} { \lambda_0(t)e^{\beta_k x_{j1} + ... + \beta_k x_{jk}}} = e^{\beta_1 x_{j1} + ... + \beta_k x_{jk}}
$$

-This defines the **proportional hazards property**

## Interpretation of the Betas
- We need to find the ratio where there is a unit increase in the covariates provided other covariates stay fixed

$$
\frac{h(t,x_1+1)}{h(t,x_1)} = \frac{h_0(t)e^{\beta_1(x_1+1)}}{h_0(t)e^{\beta_1(x_1)}} = e^{\beta_1}
$$

- We interprete   as the unit increase in log hazard per unit increase of x
- (just as log-odds in logistic regression)

## R example

- $> coxph(formula, method)$
- **formula** is linear model with a survival object as the response variable
- **method** is used to specify how to handle ties. The default is ‘efron’. Other options are ‘breslow’ and ‘exact’.
- $> cox.fit < −coxph(Surv(time, status)  x+y+z, method =‘breslow)$
- to obtain the baseline survival function $> my.survfit.object <− survfit(coxph.fit)$
